---
title: "Simulation designs, unequal visit spacing and tuning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation designs, unequal visit spacing and tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 2.9,
                      out.width = "100%", dpi = 96)
```

This vignette reproduces, on a small scale, the simulation designs of Section 6
of Hu, Wu, Zhao and Sun, and shows how `TrajChain` handles unequally and widely
spaced visits. See `vignette("TrajChain")` for an introduction to the model and
the estimands.

```{r setup}
library(TrajChain)
```

## The simulation designs

`simulate_cohort()` generates data on the unit age scale $[0, 1]$. In the
underlying population the latent variable $Z^0$ is log-normal with mean 1 and
variance 0.25, the age at death given $Z^0 = z$ is Weibull with shape 4.7 and
scale $0.85 z^{-1/4.7}$ (so that individuals with larger $Z^0$ die earlier), and
the marker is $Y^0(t) = Z^0 f(t) e_t$ with log-normal noise of mean 1 and
variance 0.05. Two shape functions are available:

```{r shapes, fig.width = 5, fig.height = 3.5, out.width = "55%"}
curve(shape_monotone(x), 0, 1, ylim = c(0.5, 1.6), lwd = 2, col = "#2a78d6",
      xlab = "t", ylab = "f(t)")
curve(shape_ushape(x), 0, 1, add = TRUE, lwd = 2, lty = 2, col = "#eb6834")
legend("topleft", c("monotone", "U-shaped"), lty = 1:2, lwd = 2,
       col = c("#2a78d6", "#eb6834"), bty = "n")
```

Entry ages follow either *random truncation*, $A^0 \sim U(-0.3, 1.9)$
independent of $Z^0$ (no birth-cohort effect), or *informative truncation*,
where $A^0 + 0.3$ given $Z^0 = z$ is Weibull with shape 3.2 and scale
$1.2 z^{-1/3.2}$ (a birth-cohort effect). Only individuals alive at entry are
enrolled (about half of the population). Each subject has a baseline visit and
up to five follow-up visits, either equally spaced with gap
$1/\lfloor 6 n^{0.16}\rfloor$ or unequally spaced with gaps
$(0.10, 0.10, 0.12, 0.10, 0.12)$; follow-up stops at death or dropout.

```{r simulate}
sim <- simulate_cohort(n = 1000, shape = "monotone", seed = 1)
sim
```

`sim_truth()` computes the true curves by numerical integration. Under random
truncation, $\mu_{Z^0}(t) = E(Z^0 \mid T^0 \ge t)$ is the pure survivor bias and
$\mu_{Z^0}(0) = 1$:

```{r truth}
sim_truth(c(0, 0.25, 0.5, 0.75, 1))
```

## Equally spaced visits

There is no birth-cohort effect and dropout is unrelated to the latent
variable, so assumption (A2') holds and the pooled estimator of
$\mu_{Z^0}(t)$ (equation 6) can be used, as in the paper.

```{r equal}
fit <- trajchain(sim$Y, sim$entry_age, sim$visit_gaps, tau0 = 0, tau1 = 1,
                 mu_method = "pooled")
plot(fit, truth = sim_truth(fit$curves$t), xlab = "t")
```

### A small Monte Carlo study

The following loop repeats the analysis for 50 data sets and summarizes the
pointwise bias and standard deviation of $\hat f$ at $t = 0, 0.1, \dots, 1$, as
in the supplementary tables of the paper (which use 1000 replicates).

```{r mc}
R <- 50
ages <- seq(0, 1, by = 0.1)
est <- t(vapply(seq_len(R), function(r) {
  s <- simulate_cohort(n = 1000, shape = "monotone", seed = r)
  f <- trajchain(s$Y, s$entry_age, s$visit_gaps, tau0 = 0, tau1 = 1,
                 mu_method = "pooled", n_grid = 101)
  predict(f, ages, type = "shape")
}, numeric(length(ages))))
round(rbind(bias = colMeans(est) - shape_monotone(ages),
            sd = apply(est, 2, sd)), 3)
```

```{r mc-plot, fig.width = 5, fig.height = 3.5, out.width = "55%"}
matplot(ages, t(est), type = "l", lty = 1, col = grDevices::adjustcolor("#2a78d6", 0.25),
        xlab = "t", ylab = "estimated f(t)")
lines(ages, colMeans(est), lwd = 2, lty = 2, col = "#e34948")
curve(shape_monotone(x), 0, 1, add = TRUE, lwd = 2)
legend("topleft", c("replicates", "pointwise mean", "truth"), bty = "n",
       col = c("#2a78d6", "#e34948", "black"), lty = c(1, 2, 1), lwd = c(1, 2, 2))
```

## Unequally spaced visits

With gaps $(5, 5, 6, 5, 6) \times 0.02$, ratios are directly estimable only for
ages $0.10$ or $0.12$ apart. `trajchain()` takes the greatest common divisor of
the gaps, $0.02$, as the grid step ($m = 50$), estimates every ratio
$\hat r(u_j, u_{j'})$ with $u_j - u_{j'} \in \{0.10, 0.12\}$, and combines them
by penalized weighted least squares (Section 5 of the paper). Ratios based on
fewer than $c^* = 0.6\,n h$ observations receive zero weight.

The penalty $\lambda$ is chosen by five-fold cross-validation over the grid
used in the paper (25 values from $10^{-6}$ to $10^{-2}$):

```{r unequal-cv, fig.width = 5, fig.height = 3.5, out.width = "55%"}
sim3 <- simulate_cohort(n = 1000, shape = "ushape", visit_gaps = "unequal", seed = 3)
cvl <- cv_lambda(sim3$Y, sim3$entry_age, sim3$visit_gaps, tau0 = 0, tau1 = 1)
cvl$lambda
plot(cvl)
```

```{r unequal-fit}
fit3 <- trajchain(sim3$Y, sim3$entry_age, sim3$visit_gaps, tau0 = 0, tau1 = 1,
                  lambda = cvl, mu_method = "pooled")
fit3
head(fit3$pairs)
plot(fit3, which = c("shape", "mean", "bias"),
     truth = sim_truth(fit3$curves$t, shape = "ushape"), xlab = "t")
```

`trajchain(..., lambda = "cv")` (the default for the penalized estimator) runs
the same cross-validation internally.

The penalized problem is normalized so that the estimate integrates to one.
By default (`constraint = "anchor"`) it is solved in closed form with the first
grid value fixed and then rescaled, which is the implementation used for the
results in the paper. `constraint = "integral"` instead solves the quadratic
program with the integral and non-negativity constraints:

```{r integral}
fit3q <- trajchain(sim3$Y, sim3$entry_age, sim3$visit_gaps, tau0 = 0, tau1 = 1,
                   lambda = cvl$lambda, constraint = "integral", mu_method = "pooled")
max(abs(fit3q$curves$shape - fit3$curves$shape))
```

## Widely spaced visits

When visits are far apart, a grid with the visit gap as its step is coarse.
Section 5 of the paper refines the grid to step $v/q$; the ratios then define
$q$ interleaved chains, which the penalty links together. In `trajchain()` this
only requires `grid_step`:

```{r wide}
simw <- simulate_cohort(n = 1000, visit_gaps = rep(0.1, 5), seed = 4)
fit_coarse <- trajchain(simw$Y, simw$entry_age, simw$visit_gaps, tau0 = 0, tau1 = 1,
                        mu_method = "pooled")
fit_fine <- trajchain(simw$Y, simw$entry_age, simw$visit_gaps, tau0 = 0, tau1 = 1,
                      grid_step = 0.05, mu_method = "pooled")
c(coarse = fit_coarse$design$m, fine = fit_fine$design$m)
plot_trajectories(list("chain, step 0.1" = fit_coarse, "penalized, step 0.05" = fit_fine),
                  which = c("shape", "mean"), truth = sim_truth(seq(0, 1, by = 0.01)),
                  xlab = "t")
```

## Informative truncation

With informative truncation, the entry age depends on $Z^0$ and
$\mu_{Z^0}(t) = E(Z^0 \mid A^0 = t, T^0 \ge t)$ combines survivor and
birth-cohort selection. The baseline estimator (equation 5), the default,
remains valid:

```{r informative}
simi <- simulate_cohort(n = 2000, truncation = "informative", seed = 5)
fiti <- trajchain(simi$Y, simi$entry_age, simi$visit_gaps, tau0 = 0, tau1 = 1)
plot(fiti, which = c("shape", "mean", "bias"),
     truth = sim_truth(fiti$curves$t, truncation = "informative"), xlab = "t")
```

## Running the full simulation study

The paper uses 1000 replicates for each design and sample size. The code below
(not run here) parallelizes over replicates and stores the curves on the
dense grid; since every replicate re-creates its data from its seed, no objects
need to be exported to the workers.

```{r full, eval = FALSE}
library(parallel)
one_replicate <- function(seed, n) {
  s <- simulate_cohort(n = n, shape = "monotone", seed = seed)
  f <- trajchain(s$Y, s$entry_age, s$visit_gaps, tau0 = 0, tau1 = 1, mu_method = "pooled")
  f$curves[, c("shape", "mean_pooled", "bias_pooled")]
}
res <- mclapply(1:1000, one_replicate, n = 1000, mc.cores = detectCores() - 1)
shape_mat <- t(vapply(res, function(r) r$shape, numeric(801)))
```
