---
title: "Getting started with TrajChain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with TrajChain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 2.9,
                      out.width = "100%", dpi = 96)
```

`TrajChain` estimates population-level biomarker trajectories from longitudinal
cohort studies in which individuals enter at different ages and are followed
until death or dropout. It implements the nonparametric framework of Hu, Wu,
Zhao and Sun (*Nonparametric estimation of marker trajectories in the presence
of death and delayed entry*, manuscript under revision at *Biometrika*). This
vignette explains the estimands and walks through a complete analysis of a
simulated cohort.

```{r setup}
library(TrajChain)
```

## Why naive age curves are biased

A common way to summarize how a biomarker changes with age is to smooth the
observed measurements against age. In a cohort that enrolls a cross-sectional
sample and follows it over time, such a curve mixes three things:

* the **age-related change** that every individual undergoes;
* **survivor bias**: at older ages only survivors are observed, and if death is
  associated with the biomarker through unobserved factors, survivors are a
  selected group;
* **birth-cohort effects**: individuals observed at older ages were born
  earlier. Delayed entry (left truncation) means that the cohorts are sampled
  differently at each age, so differences between birth cohorts leak into the
  age curve.

## Model and estimands

Let $\tau_0$ be the starting age of the trajectory (age 30 in the ABC-DS
application of the paper); the target population consists of individuals alive
at $\tau_0$. For such an individual let $Y^0(t)$ be the marker at age $t$, $T^0$
the age at death and $Z^0>0$ an unobserved latent variable. The model is

$$E\{Y^0(t) \mid Z^0, T^0 = s\} = Z^0 f(t), \qquad \tau_0 \le t \le s,$$

with $\int_{\tau_0}^{\tau_1} f(t)\,dt = 1$. The **shape function** $f$ is the
common age-related pattern; $Z^0$ captures individual heterogeneity and may be
associated with survival, with the age at study entry (birth cohort) and with
dropout, in ways that are left unspecified. Neither survival times nor the
reason for dropout need to be observed. Two conditional independence
assumptions are needed:

* **(A1)** the age at entry $A^0$ is independent of the marker process given
  $Z^0$ and the age at death, so birth cohorts may differ only through $Z^0$;
* **(A2)** the dropout index $C$ (the last follow-up visit before loss to
  follow-up for reasons other than death) is independent of the age at death
  and the marker process given the entry age and $Z$.

Dropout that depends on the marker values themselves (for example, sicker
participants leaving the study) violates (A2).

For individuals entering the study at age $t$ (alive at entry), the mean of the
observed marker is

$$E\{Y(A) \mid A = t\} = \mu_{Z^0}(t)\, f(t), \qquad
\mu_{Z^0}(t) = E(Z^0 \mid A^0 = t, T^0 \ge t),$$

so a naive Nadaraya--Watson curve estimates the shape distorted by
$\mu_{Z^0}(t)$. `TrajChain` reports

| Quantity | Formula | Column in `fit$curves` |
|---|---|---|
| Shape function | $f(t)$ | `shape` |
| Unconditional mean | $\mu_{Z^0}(\tau_0)\, f(t)$ | `mean_*` |
| Birth-cohort-survivor bias | $\mu_{Z^0}(t)/\mu_{Z^0}(\tau_0)$ | `bias_*` |
| Latent mean among survivors | $\mu_{Z^0}(t)$ | `mu_*` |
| Simple kernel smoothing | $E\{Y(A) \mid A = t\}$ (baseline) or all visits (pooled) | `nw_*` |

The unconditional mean is the mean trajectory in the absence of death; with
birth-cohort effects it refers to the subgroup whose starting age coincides
with the study baseline. The bias curve equals one at $\tau_0$; values below one
mean that survivors entering at age $t$ have a lower latent level than the
reference group, i.e. lower marker levels than the reference group would have
at the same age.

**How $f$ is identified.** For two ages $v$ apart, under (A1)--(A2) the
nuisance factors cancel in the ratio of means of consecutive measurements, so
$r(t, t-v) = f(t)/f(t-v)$
is estimated by a kernel ratio estimator (`ratio_estimate()`). With equally
spaced visits these ratios are chained along the grid
$u_j = \tau_0 + (j - 0.5)v$ and normalized to integrate to one (*ratio
chaining*). With widely or unequally spaced visits, `trajchain()` matches all
estimable ratios by penalized least squares.

## The data

`simcohort` is a simulated cohort in long format (one row per measurement): 600
subjects in two groups, a baseline visit and up to two follow-up visits every 16
months, with deaths, dropout and informative delayed entry.

```{r data}
head(simcohort)
table(simcohort$group, simcohort$visit)
```

`trajchain()` needs a subject-by-visit matrix (`NA` for visits that did not take
place), the entry ages and the scheduled time between visits.
`visit_matrix()` builds the matrix from long data:

```{r matrix}
dA <- subset(simcohort, group == "A")
YA <- visit_matrix(dA, id = "id", visit = "visit", value = "marker")
AA <- dA$entry_age[match(rownames(YA), dA$id)]
head(YA, 4)
```

Visits are scheduled every 16 months, so the grid step is $v = 16/12$ years. We
study ages 30 to 59.33, i.e. $m = 22$ grid points, as in the paper's analysis.

## Fitting the model

```{r fit}
fitA <- trajchain(YA, AA, visit_gaps = 16 / 12, tau0 = 30, m = 22)
fitA
```

The defaults follow the paper: ratio bandwidth
$h = \hat\sigma_A n^{-1/6}$, mean bandwidth $h' = 2.34\,\hat\sigma_A n^{-1/5}$,
Epanechnikov kernel, and the baseline-measurement estimator of
$\mu_{Z^0}(t)$, which remains valid in the presence of birth-cohort effects.
Because `simcohort` is simulated, we can add the true curves (dashed) to the
plot:

```{r truth}
tau <- 22 * 16 / 12
truth_age <- function(shape) {
  s <- seq(0, 1, length.out = 201)
  tr <- sim_truth(s, shape = shape, truncation = "informative")
  data.frame(t = 30 + tau * s, shape = tr$shape / tau, mean = tr$mean, bias = tr$bias)
}
plot(fitA, truth = truth_age("monotone"))
```

The simple kernel smoothing curve (third panel) rises until about age 52 and
then declines, although the underlying shape keeps increasing: survivors at
older ages have lower latent levels, which is what the decreasing
birth-cohort-survivor bias (fourth panel) quantifies. The estimated shape and
unconditional mean recover the increasing trend. The estimated bias shows a
bump in the thirties that the true curve does not have; the bootstrap bands
below show that this part of the curve is estimated with considerable
uncertainty. The bias and the unconditional mean are anchored at age 30, and few
subjects enter the study close to that age.

Estimates at particular ages are available from `summary()` and `predict()`:

```{r summary}
summary(fitA, ages = c(30, 40, 50, 59))$table
predict(fitA, c(35, 45, 55), type = "mean")
```

## Bandwidth: rule of thumb and cross-validation

The constant $\kappa$ in $h = \kappa\,\hat\sigma_A n^{-1/6}$ can be chosen by
five-fold cross-validation with `cv_bandwidth()`. Its criterion compares, in
held-out subjects, consecutive measurements with the ratio predicted from the
training subjects. With small samples, averaging over several random partitions
(`n_repeats`) reduces partition noise.

```{r cv, fig.width = 5, fig.height = 3.5, out.width = "60%"}
cvA <- cv_bandwidth(YA, AA, visit_gaps = 16 / 12, tau0 = 30, m = 22, n_repeats = 5)
cvA
plot(cvA)
```

Here the criterion is very flat for $\kappa \ge 1$: the data hardly discriminate
between these bandwidths. A useful practice, as in the paper, is to report the
rule-of-thumb fit ($\kappa = 1$) and check the sensitivity of the conclusions to
$\kappa$:

```{r sensitivity}
fits_kappa <- lapply(c(0.75, 1, 1.5, cvA$constant), function(k) {
  trajchain(YA, AA, 16 / 12, tau0 = 30, m = 22, bandwidth = k * bw_rot(AA))
})
names(fits_kappa) <- paste("kappa =", c(0.75, 1, 1.5, cvA$constant))
plot_trajectories(fits_kappa, which = c("shape", "mean", "bias"))
```

The shape function is stable across bandwidths. The unconditional mean and the
bias curve are anchored at the starting age through
$\hat\mu_{Z^0}(\tau_0)$, which divides the kernel smoother at $\tau_0$ by
$\hat f(\tau_0)$, a value extrapolated from the first grid points; they are
therefore more sensitive to the bandwidth, as observed in the data analysis of
the paper.

## Bootstrap confidence bands

`trajchain_boot()` resamples subjects with replacement and refits with the
bandwidths held fixed, giving pointwise standard errors and percentile
confidence bands for every curve.

```{r boot}
bootA <- trajchain_boot(fitA, B = 200, seed = 2026)
summary(bootA, ages = c(35, 45, 55), which = c("shape", "bias"))
plot(bootA)
```

The bias curve is fixed at one at age 30 by construction, so its band has zero
width there. This reflects the normalization rather than precise estimation at
age 30: the uncertainty about $\hat\mu_{Z^0}(30)$ shows up in the width of the
bias band at later ages and in the band of the unconditional mean.

## Comparing groups

Analyses stratified by a grouping variable (sex in the paper) are run
separately and displayed together with `plot_trajectories()`:

```{r groups}
fit_group <- function(g) {
  d <- subset(simcohort, group == g)
  Y <- visit_matrix(d, id = "id", visit = "visit", value = "marker")
  A <- d$entry_age[match(rownames(Y), d$id)]
  trajchain(Y, A, visit_gaps = 16 / 12, tau0 = 30, m = 22)
}
bootB <- trajchain_boot(fit_group("B"), B = 200, seed = 2027)
plot_trajectories(list("Group A" = bootA, "Group B" = bootB))
```

## Which estimator of the mean?

Both estimators of $\mu_{Z^0}(t)$ are computed in every fit:

* `mu_method = "baseline"` (equation 5 of the paper) smooths the baseline
  measurements only. It is valid under assumptions (A1)--(A2), which allow the
  entry age, death and dropout to depend on the latent variable, and therefore
  allows birth-cohort effects. This is the default and was used for the ABC-DS
  analysis.
* `mu_method = "pooled"` (equation 6) pools all observed visits. It requires
  dropout to be independent of $(T, Y, Z)$ given the entry age (assumption
  A2') and approximates $\mu_{Z^0}(t)$ when the visit gaps are small; with
  birth-cohort effects and wider gaps it mixes subjects who entered at
  different ages. It was used in the simulation studies of the paper, where
  there is no birth-cohort effect.

```{r pooled}
head(fitA$curves[, c("t", "bias_baseline", "bias_pooled")], 3)
plot(fitA, which = c("mean", "bias"), mu_method = "pooled")
```

## Practical notes

* **Scheduled visit times.** Visit $l$ of subject $i$ is placed at
  $A_i + v_1 + \dots + v_l$, the scheduled time. Small deviations of the actual
  visit dates are ignored, as in the paper.
* **Monotone dropout.** The method assumes that a subject is observed at visits
  $0, 1, \dots, C$ and not afterwards. If a visit is missing but a later one is
  observed, `trajchain()` warns and uses only pairs of consecutive observed
  visits.
* **Undefined ratios.** If some grid age has no pair of consecutive visits within
  one bandwidth, the ratio estimator is undefined there. The chain estimator
  then stops with an informative error (increase the bandwidth or narrow
  $[\tau_0, \tau_1]$); the penalized estimator drops such ratios.
* **Choice of $\tau_0$.** The unconditional mean and the bias curve are
  anchored at $\tau_0$, so $\tau_0$ should lie well within the range of entry
  ages; `trajchain()` warns when fewer than 10 subjects enter within one
  bandwidth $h'$ of $\tau_0$.
* **Unequally or widely spaced visits.** Give the vector of gaps in
  `visit_gaps` (or a finer `grid_step`); `trajchain()` then uses the penalized
  estimator and chooses the penalty by cross-validation. See
  `vignette("simulation", package = "TrajChain")`.

## Reference

Hu, P., Wu, Y., Zhao, Y. and Sun, Y. Nonparametric estimation of marker
trajectories in the presence of death and delayed entry. Manuscript under
revision at *Biometrika*.
